Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-08-22 Manuscript/01-Intro.Rmd
Run pipeline

Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when Cleveland [@Cleveland1979] proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed. [@Wood2020] With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classic statistical models. However, these models have a fatal flaw, the lack of transparency and interpretability, and hence are referred to as "black box" models [@Breiman2001]. Given that the primary objective of biomedical studies is to understand the etiology of diseases, "Black box" models are less helpful as they can hardly explain how the change of biomarkers is associated with the disease.

Semi-parametric regression models are a more appropriate alternative to the "black-box" models, thanks to their balancing between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be expressed as the summation of smooth functions, $f_j, j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
$$
where $g^{-1}$ is the inverse function of a smooth monotonic link function. The smooth functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classic GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

Many modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see Mallick and Yi [@Mallick2013] for review. These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients: Ravikumar et al. [@Ravikumar2009] extended the grouped lasso [@Yuan2006] to additive models; Huang et al. [@Huang2010] further developed adaptive grouped lasso for additive models; Wang et al. [@Wang2007] and Xue [@Xue2009] respectively applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed, where the application on AMs are by-products. Scheipl et al. [@Scheipl2012] proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma, specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. Bai et al. [@Bai2020] was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs[@Bai2021]. Regardless of the nice statistical properties, a practical and pivotal question, if non-linear effects exist, yet to be addressed by these methods. Besides Scheipl et al. [@Scheipl2012], most proposed methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for within group selection. In addition, a common criticism received by these methods is that the basis function coefficients tend to be overly penalized due to the sparsity assumption directly applied on the basis function coefficients.[@Scheipl2013] It would be of special interest to develop a fast yet flexible generalized additive model.

>>>>

Many modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see Mallick and Yi [@Mallick2013] for review. These models are built upon a restrictive and unrealistic assumption, linearity. In classical statistical modelling, many modeling strategies and models are proposed to relax the linearity assumption with various degrees of complexity. For example, variable categorization is a simple and common practice in epidemiology, but suffers from power and interpretation issues. More complex models to address nonlinear effects include random forest and other so-called "black box" models [@Breiman2001]. These models are not particularly attractive in biomedical studies, as they lack the transparency to understand the etiology of diseases. In addition, how to generalized these models to the high-dimensional setting remains unclear.

<!-- Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when Cleveland [@Cleveland1979] proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed. [@Wood2020] With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classic statistical models.  -->

<!-- However, these models have a fatal flaw, the lack of transparency and interpretability, and hence are referred to as "black box" models [@Breiman2001]. Given that the primary objective of biomedical studies is to understand , "Black box" models are less helpful as they can hardly explain how the change of biomarkers is associated with the disease. -->

Semi-parametric regression models are an appropriate alternative to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be expressed as the summation of smooth functions, $f_j, j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
$$
where $g^{-1}$ is the inverse function of a monotonic link function. The smooth functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

There exists some proposals to generalized the classical GAM to accommodate high-dimensional applications. The regularized models that are primarily designed for linear models are used to fit GAMs by accounting for the group structure introduced when expanding smoothing functions.
<!-- These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients:  -->
Ravikumar et al. [@Ravikumar2009] extended the grouped lasso [@Yuan2006] to additive models; Huang et al. [@Huang2010] further developed adaptive grouped lasso for additive models; Wang et al. [@Wang2007] and Xue [@Xue2009] respectively applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed, where the application on AMs are by-products. Scheipl et al. [@Scheipl2012] proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma, specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. Bai et al. [@Bai2020] was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs[@Bai2021].

TODO: emphasize it doesn't accounting for smoothing penalty, and possibly lead to overshinkage, resulting wrong estimation of the smoothing function and prediction.

Regardless of the nice statistical properties, a practical and pivotal question, if non-linear effects exist, yet to be addressed by these methods. Besides Scheipl et al. [@Scheipl2012], most proposed methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for within group selection. In addition, a common criticism received by these methods is that the basis function coefficients tend to be overly penalized due to the sparsity assumption directly applied on the basis function coefficients.[@Scheipl2013] It would be of special interest to develop a fast yet flexible generalized additive model.
